`Total-evidence' Bayesian estimation of phylogeny, divergence times and fossil ages

### 2016-04-30 response-to-reviewers.tex

Please give the prior on the morphological substitution rate, and for relaxed-clock models, the prior on the rate variation parameter (I am assuming you are using log-normal clock, so prior on log-variance).

\response{We have added the prior information for the morphological substitution rate and log standard deviation of the relaxed clock.}

p.8 paragraph starting on line 43. Do you mean that with the particular implementation of marginal likelihood calculation in BEAST you cannot calculate the BF? Or is this a general problem with any implementation of stepping-stones and this model?

\response{One can calculate BF either in the way described here or via stepping stone approaches. Stepping stone approaches have been used for the coalescent, where the number of tips in a tree is fixed. Here, the number of trees is treated as part of the data, and while a stepping stone approach may be still appropriate, we did not test it for this scenario. This point is highlighted in ``We want to note that the stepping stone sampling approach \cite{Xie2011}
 to calculate Bayes factors would not have been directly applicable in our case:  In stepping stone sampling the $D_\mathcal{T}$ is treated as part of the model, not part of the data. However, using a birth-death model, the sampling times are part of the data. The approach is valid when choosing a  coalescent tree prior, as in that case sampling times are conditioned upon (and thus can be seen as part of the model) rather than being modelled (and thus are a realisation of the model which means they are data). It is not clear if the stepping stone approach can be directly applied for models with number of tips being part of the data.
In general, even if stepping stone approaches are appropriate, we recommend inspection of $P(H_1|M)$ to ensure that the prior on the hypothesis to be tested is sensible. Such an investigation reveals if the cause of a high (or low) Bayes factor is due to the prior or due to signal in the data.
'' }

@@

I am troubled about the sensitivity of the BF to the prior in this case. Can the authors give some clear, easy to follow, recommendations for biologists that may wish to use the method in their own works?

\response{Unfortunately we can't. This issue will need further attention in future work.}

Another issue is that when the model is changed, the BF change. The authors should point out to the reader that rejecting the stratigraphic estimate of a fossil with BF does not prove that the stratigraphic age is wrong. If the morphological or clock model are misspecified (i.e. they are  bad description of the data), then I would expect the BF to fail. For example, if a lineage has shown explosively fast rates of evolution, then I would expect the Bayesian phylogenetic estimates of the fossil age to contradict the stratigraphic range if a strict clock is imposed.

\response{We added: A BF $>>$ 1 indicates strong support for $H_1$, given the model $M$ is appropriate for the considered data.}

@@

One concern I have with this method, which perhaps doesn't apply to these datasets, is how it handles situations where there is rampant morphological convergence across species in the dataset. For example, previous analyses have shown that large morphological datasets group together mammals with convergent ecotypes (e.g., ant-eating, tree-dwelling, hoofed; Springer et a. 2013 Science) and squamates without legs (e.g., snakes, amphisbaenians, anniellids; Gauthier et al. 2012 Bulletin of the Peabody Museum of Natural History), respectively, in sharp contrast to both molecular data and, frequently, the fossil record. Presumably it should lead to highly inaccurate phylogenetic age estimates, though this might also be a potential application of the method (i.e., highly inaccurate dates may be suggestive of homoplasy appropriately grouping taxa together).

This raises a further question of whether this method can be used in conjunction with molecular data to correct for morphological homoplasy. Beyond this, perhaps one could test how the molecular dataset can influence the phylogenetic age of these fossils, and could provide further insights regarding whether divergence times derived from molecular clocks are accurate. I'm not suggesting more analyses, but perhaps further discussion of these issues.

\response{We added a paragraph on the appropriateness of a morphological clock: paragraph 1 in ``Rates of evolution and relaxed morphological clocks
''}

### 2016-04-30 fossilDating-revision.tex

\section*{Methods}
%TSrev
Gavryushkina {\it et al} \cite{gavryushkina2015bayesian} described a ``total-evidence'' approach implemented in BEAST2 \citep{Beast2} for phylogenetic estimation of time-trees that employs both morphological data from fossils and extant data and molecular sequence data as equal partners under the rule of probability for estimating a time-tree.  An equivalent method \citep{zhang2016} is introduced within MrBayes \cite{ronquist2012mrbayes}. 
The model of time-tree phylogeny employed is the so-called fossilized birth-death process \cite{Heath2014}, which forms a prior probability distribution on the space of sampled-ancestor trees \cite{Gavr2013}.

@@

%\alexei{%
We used a data set originally published by \cite{ksepka2012} consisting of morphological data from fossil and living penguin 
species. 
We used the same subset of the morphological data as in \cite{gavryushkina2015bayesian} but we did not use the molecular sequence data from the living species. 
The morphological data matrix we used contains 36 fossil species, 19 extant species and 202 characters (ranging from binary to k=7). 
The majority of these characters (\textgreater 95\%) have fewer than four states and 48 of the binary characters were encoded as presence/absence. 
The fossil ages ranged from TODO.
As did \cite{gavryushkina2015bayesian}, we treat 34 characters that were ordered in \cite{ksepka2012} as unordered. 
See \cite{gavryushkina2015bayesian} for further details of data selection.
%}%

@@

%TSrev
We applied two phylogenetic models to the penguin data set, \Mstrict{} and \Mrelaxed{}, and we applied \Mstrict{} to the canid dataset. \Mstrict{} assumed a strict morphological clock ($\mu$) and no gamma-distributed rate heterogeneity among sites  \cite{Lewis2001}. Model \Mrelaxed{} \citep{gavryushkina2015bayesian} partitioned the alignment into  partitions (six for the penguins), with the $i$'th partition containing all characters that had $k=i+1$ character states across the sampled taxa. Model \Mrelaxed{} also uses a uncorrelated lognormally-distributed relaxed molecular clock \cite{Drummond2006} with parameters $\mu$ and $S$ for the mean rate and log standard deviation of the rates, and an additional parameter for gamma-distributed rate variation across sites \cite{yang:1994ma}. 
%TODO how was this additional parameter set?
We followed \cite{gavryushkina2015bayesian} in having a broad LogNormal(-5.5, 2) prior on $\mu$ for all analyses and a Gamma($\alpha=0.5396$, $\beta=0.3819$) prior on $S$ for the relaxed clock analyses.
%TSref TODO Alexei please check if the following sentence is correct.
For the parameters of the fossilized birth-death model tree prior, 
we specify the prior below in the section ``Computing the phylogenetic evidence for an age range''.
%we  also followed \cite{gavryushkina2015bayesian}, except with an upper limit on the origin of $T= 80$ for the canids.

>>>>

\section*{Methods}
%TSrev
Gavryushkina {\it et al} \cite{gavryushkina2015bayesian} described a ``total-evidence'' approach implemented in BEAST2 \citep{Beast2} for phylogenetic estimation of time-trees that employs both morphological data from fossils and extant data and molecular sequence data as equal partners under the rule of probability for estimating a time-tree.  An equivalent method \citep{zhang2016} is introduced within MrBayes \cite{ronquist2012mrbayes}. 
The model of time-tree phylogeny employed is the so-called fossilized birth-death process \cite{Heath2014}, which forms a prior probability distribution on the space of sampled-ancestor trees \cite{gavr2014,Gavr2013}.

@@

%\alexei{%
We used a data set originally published by \cite{ksepka2012} consisting of morphological data from fossil and living penguin 
species. 
We used the same subset of the morphological data as in \cite{gavryushkina2015bayesian} but we did not use the molecular sequence data from the living species. 
The morphological data matrix we used contains 36 fossil species, 19 extant species and 202 characters (ranging from binary to k=7). 
The majority of these characters (\textgreater 95\%) have fewer than four states and 48 of the binary characters were encoded as presence/absence. 
The fossil ages ranged from 4.5 to 61.3 Myr.
As did \cite{gavryushkina2015bayesian}, we treat 34 characters that were ordered in \cite{ksepka2012} as unordered. 
See \cite{gavryushkina2015bayesian} for further details of data selection.
%}%

@@

%TSrev
We applied two phylogenetic models to the penguin data set, \Mstrict{} and \Mrelaxed{}, and we applied \Mstrict{} to the canid dataset. \Mstrict{} assumed a strict morphological clock ($\mu$) and no gamma-distributed rate heterogeneity among sites  \cite{Lewis2001}. Model \Mrelaxed{} \citep{gavryushkina2015bayesian} partitioned the alignment into  partitions (six for the penguins), with the $i$'th partition containing all characters that had $k=i+1$ character states across the sampled taxa. Model \Mrelaxed{} also uses a uncorrelated lognormally-distributed relaxed molecular clock \cite{Drummond2006} with parameters $\mu$ and $S$ for the mean rate and log standard deviation of the rates, and an additional parameter for gamma-distributed rate variation across sites \cite{yang:1994ma}. 
%TODO how was this additional parameter set?
We followed \cite{gavryushkina2015bayesian} in having a broad LogNormal(-5.5, 2) prior on $\mu$ for all analyses and a Gamma($\alpha=0.5396$, $\beta=0.3819$) prior on $S$ for the relaxed clock analyses.
%TSref TODO Alexei please check if the following sentence is correct.
For the penguin analyses, the parameters of the fossilized birth-death model tree prior  were specified as described in the section ``Computing the phylogenetic evidence for an age range''. 
For the canid analysis we used a more standard parametrisation: origin $T$ with Uniform(0,120) prior, diversification rate $d$ with LogNormal(-3.5, 1.5), turnover and sampling proportion with Uniform(0,1) priors.
